---
title: "The Rent is Too Damn High"
author: "Ethan Jantz & Alex Kwan"
date: "4/17/2021"
output:
  html_document: 
    code_folding: hide
---

```{r setup,  message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(gganimate) # Used to create an animated plot
# install.packages('gifski') necessary for animation to compile
library(ggrepel) # Used to create cleveland dot plot
library(here)
library(patchwork) # Used to arrange plots as one image
library(sf) # Used to manipulate the spatial data for maps
library(tidycensus) # Used to pull census data 
library(tidyverse)
library(tigris) # Used to pull spatial features

# Ingests and cleans the HUD data
# Creates the object hud_data_raw in env
# Source: https://www.huduser.gov/portal/datasets/fmr.html
# Methodology: https://www.huduser.gov/portal/sites/default/files/pdf/fmr-overview.pdf
# NOTE: some years have more ZIP codes than others in this data
source(here("ingest_hud_data.R"))

# Zillow Observed Rent Index (ZORI) data
# Source: https://www.zillow.com/research/data/
# NOTE: Not every ZIP code in Chicago is included in this data
zori_raw <- read_csv(here("Data", "ZORI_AllHomesPlusMultifamily_SSA_ZIP.csv")) %>%
  filter(RegionName %in% hud_data_raw$ZIP) # 92 ZIP Codes
```

# Background

Housing is a [social determinant of health](https://health.gov/healthypeople/objectives-and-data/social-determinants-health). Affordable housing is related to better [health outcomes for tenants](https://www.rupco.org/wp-content/uploads/pdfs/The-Impacts-of-Affordable-Housing-on-Health-CenterforHousingPolicy-Maqbool.etal.pdf), freeing money up for better nutrition, daycare and healthcare, and other necesities for individual wellbeing. Evictions are related to [poor health outcomes](https://www.bu.edu/sph/news/articles/2018/the-hidden-health-crisis-of-eviction/). Housing instability has been found to lead to higher stress levels, increased rates of depression and anxiety, and increased likelihood of living in lower quality housing or homelessness. Because of this, it is important to understand current metrics for what constitutes affordable housing. This project explores housing affordability through federal government metrics used to calculate housing subsidies known as Fair Market Rent (FMR).

Fair Market Rent is determined by a combination of a census survey of rents paid by recent movers, rent inflation adjustment, and forecasted expected growth in the rental housing market. FMR includes contract rent and utilities, ensuring that a full cost of base living necessities are captured in the metric. FMR calculations are made using American Community Survey from two years prior to the calculation and are primarily used to set public housing flat rates and Section 8 housing choice voucher payment standards. [Housing choice voucher recipients](https://www.huduser.gov/portal/sites/default/files/pdf/fmr-overview.pdf) and [public housing tenants](https://www.huduser.gov/periodicals/ushmc/spring95/spring95.html) make up approximately 3 million, or 2.5%, of United States households. Because of this scope, HUD FMR calculations can serve as a lense into the effectiveness of housing markets across the country in addressing issues of housing provision. 

# Research Questions

- Is the U.S. Department of Housing and Urban Development’s Fair Market Rent calculations representative of actual rents in an area?
- What have metrics for housing subsidies looked like over time?
- If there is a difference between actual rents and Fair Market Rents as determined by HUD, what data can we use to explain those differences?

# Data Plan

To perform this analysis, we needed data from three sources: 

* Fair Market Rent (FMR) data from the United States Department of Housing and Urban Development (HUD)

* Observed rent data provided by Zillow

* Socio-economic data provided by the United States Census in their American Community Survey data product

All of these data needed to be available at the same geographic level, and thankfully that provided no challenge as all of these data products are aggregated to the ZIP code level and the Metropolitan Statistical Area level. For this project we focused our efforts on ZIP codes within the City of Chicago. HUD FMR data is available through the [HUD Policy Development & Research Website](https://www.huduser.gov/portal/datasets/fmr.html)

# Analysis Plan

Zillow's ZORI data provides average rents across all units, **regardless of the number of bedrooms**. We need some method to compare ZORI values with the values provided by HUD. This could take the form of finding the best fitting bedroom size or collapsing the HUD bedroom values into one average value. We can try both and see what works best.

```{r }

### The final dataset should have no missing values for a ZIP code in any year. That means that the ZIP codes should have values for all years in the the hud and zori datasets.

### --------------
### hud data
###

hud_data <- hud_data_raw %>%
  # Create a variable to map to the ZORI values
  mutate(br_avg = (br_0 + br_1 + br_2 + br_3 + br_4) / 5) %>% 
  select(!c(br_0:br_4)) %>%
  # saw some duplicates on the year-zip level
  distinct()

### ------------
### zori data
###

zori_staging <- zori_raw %>%
  filter(RegionName %in% hud_data_raw$ZIP) %>% # 92 ZIP Codes
  pivot_longer(cols = "2014-01":"2020-12",
               names_to = c("year","month"),
               names_sep = "-",
               values_to = "rent",
               names_transform = list(year = as.double)) %>%
  select("RegionName","year","month","rent") 

# quite a bit of missing data, ignoring issue for now
# zori_staging %>%
#   group_by(RegionName, year) %>%
#   summarize(rent_missing_prop = sum(is.na(rent))/12)

zori <- zori_staging %>%
  group_by(RegionName, year) %>%
  summarize(rent_avg = mean(rent, na.rm = TRUE)) %>%
  ungroup()

rm(zori_staging)

### -----------
### combined
###

data <- zori %>%
  left_join(hud_data, by = c("RegionName" = "ZIP", "year" = "year")) %>%
  rename(hud_avg = br_avg, zori_avg = rent_avg)
```

# Plots

Once we have both our data fully cleaned we'll want to do some basic visualizations of the data. I see this taking the form of a distribution plot across ZIP codes by year, a series of maps by year, and trend lines of ZORI values and FMR values.

```{r }
### -----------
### Trend plot
###
# Shows the parallel aggregate trend lines between observed rents and HUD's FMR values 
# average across ZIP codes observed in data. We need to explain what this is,
# and why it's the case. It's because FMR is derived from actual rents, and
# the process for determining FMR excludes the newest housing stock (buildings
# constructed in the past 2 years). 
data %>%
  group_by(year) %>%
  summarize(ZORI = mean(zori_avg, na.rm = T),
            HUD = mean(hud_avg, na.rm = T)) %>%
  pivot_longer(cols = c(ZORI:HUD),
               names_to = "index",
               values_to = "rent") %>%
  ggplot(aes(x = year, y = rent, color = index)) +
  geom_line(size = 2, lineend = "round") +
  labs(title = "Average Rents",
       subtitle = "HUD Fair Market Rent Average and Zillow Observed Rent Index Average",
       x = "Year",
       y = "Rent",
       color = "Source") +
  theme_minimal()
```

# Exploring Change in Rent Valuations

There may be a spatial component to these FMR valuations compared with observed rent. This chunk maps the data to the ZIP code geometry provided by `tigris`.

```{r }
### -----------------------------
### Year-over-Year difference map
### Prepare data

# Transforming to this CRS makes the map look "normal"
crs <- 3435 # coordinate reference system: IL NAD83 East

# Get Cook County geometry to put zctas in context
cook_county <- counties(state = "IL", cb = T, progress_bar = F) %>%
  filter(NAME == "Cook") %>% # Chicago + suburbs are almost entirely within Cook County
  select(NAME, geometry) %>%
  st_transform(crs)

# Create filter for ZIP geom pull
starts_with <- data %>% 
  pull(RegionName) %>% # Vectorize the column
  str_extract(pattern = "[0-9]{2}") %>%
  unique() # c("53", "60") 

### Get ZCTA geometry
zcta <- zctas(cb = T, starts_with = starts_with, progress_bar = F) %>%
  select(GEOID10, geometry) %>%
  st_transform(crs) %>% # Without this transformation to ILNAD83 the map looks weird
  st_filter(cook_county) # There are ZIP codes that fall outside of Cook county
                         # that make the map less easy to digest. This removes them
```

The following map shows that the difference between ZORI and HUD FMR values has been shrinking, and in some places reversing. 

```{r }
### -----------------------------
### Year-over-Year difference map
### Animation

map_anim <- data %>%
  # NOTE: Because of the way this is calculated we don't 
  # have any indication if the change is due to a change in 
  # ZORI or HUD values
  mutate(diff = zori_avg - hud_avg) %>% 
  left_join(zcta, by = c("RegionName" = "GEOID10")) %>%
  ggplot(aes(geometry = geometry)) +
  geom_sf(data = cook_county) + # County boundary
  geom_sf(aes(fill = diff)) + # ZCTA map
  labs(title = "Year: {current_frame}", # Adds the year value to the title through interpolation
       caption = "In some ZIP codes, the difference between ZORI and HUD FMR has changed drastically",
       fill = "ZORI - HUD") + 
  scale_fill_viridis_c() +
  theme_minimal() +
  transition_manual(year) # transition_time doesn't work well with sf for some reason

animate(map_anim, fps = 1, nframes = 6)
```


The following distribution plots help us understand that this is an effect of HUD FMR increasing, not ZORI decreasing. What does this mean for housing affordability? This could mean that HUD voucher recipients are being provided a wider range of places to move to as HUD FMR begins to match average rents.

```{r }
### --------------
### ZIP/Rent distribution plot
###

# NOTE: These values aren't inflation adjusted, which would be important for a more in-depth analysis. 
# However, since $1000 in 2014 is equivalent to $1118.87 today we can ignore it without
# running into any major methodological issues.

# NOTE: This could be animated too
dist_anim <- data %>% 
  group_by(RegionName, year) %>%
  summarize(ZORI = mean(zori_avg, na.rm = T),
         HUD = mean(hud_avg, na.rm = T)) %>%
  ggplot() +
  geom_density(aes(x = ZORI), fill = "blue", alpha = .5) +
  geom_density(aes(x = HUD), fill = "red", alpha = .5) +
  labs(title = "Distribution of Rent Values by ZIP Code, {current_frame}",
       subtitle = "blue is ZORI, red is HUD",
       x = "Rent", 
       y = "Density") +
  transition_manual(year)

animate(dist_anim, fps = 1, nframes = 6)
```

# Incorporating Census Data

We'll want to take the above analysis and tie it back to the policy implications. We know HUD FMR data is used to guide housing subsidies. What does that mean in terms of the rental housing market, tenant precarity, and larger trends of income and housing inequality? If we have time I think it would be good to bring in some census data at the ZCTA level and looking to see if there is any association between demographics and housing affordability using this data.

```{r}
# This entire section could be cleaned up, but it works.

source(here("ingest_census_data.R"))

zips <- data %>%
  pull(RegionName) %>%
  unique()

acs_2014 <- acs5_pull(2014, zips) %>%
  # Including zips in the function call is supposed to prevent us needing
  # to filter by zip, but it's not working.
  mutate(ZIP = str_extract(NAME, "[0-9]{5}?")) %>%
  filter(ZIP %in% zips)

acs_2019 <- acs5_pull(2019, zips) %>%
  mutate(ZIP = str_extract(NAME, "[0-9]{5}?")) %>%
  filter(ZIP %in% zips)

# TODO: This can be put into one dataset so that we can
# analyze change over time
data_2014 <- data %>%
  filter(year == 2014) %>%
  left_join(acs_2014, by = c("RegionName" = "ZIP", "year"))

data_2019 <- data %>%
  filter(year == 2019) %>%
  left_join(acs_2019, by = c("RegionName" = "ZIP", "year"))
```

## Plotting Associations

```{r}
### -------------
### coor plot
### ZORI and hh_income

plot1 <- data_2014 %>%
  ggplot(aes(x = median_hh_income, y = zori_avg)) +
  geom_jitter() +
  geom_smooth(method = "lm") +
  labs(title = "Observed Rents and Household Income, 2014",
      subtitle = "for select ZIP Codes",
      x = "Median Household Income",
      y = "Zillow Observed Rent Index Value")

plot2 <- data_2019 %>%
  ggplot(aes(x = median_hh_income, y = zori_avg)) +
  geom_jitter() +
  geom_smooth(method = "lm") +
  labs(title = "Observed Rents and Household Income, 2019",
      subtitle = "for select ZIP Codes",
      x = "Median Household Income",
      y = "Zillow Observed Rent Index Value")

plot1 + plot2
```

```{r }
### ------------------
### Cleveland Dot Plot
### Animated

neighborhoods <-
  list(
    "60045" = "Lake County",
    "60602" = "Downtown",
    "60610" = "Near Northside, Cabrini Green",
    "60126" = "Elmhurst",
    "60642" = "River West",
    "60647" = "Logan Square",
    "60616" = "Chinatown, Armour Square",
    "60525" = "La Grange",
    "60612" = "East Garfield Park",
    "60608" = "Pilsen"
  )

cleveland_data <-
  data %>%
  # can't do functions in label assignments so gotta do it here
  mutate(diff = (zori_avg - hud_avg)/hud_avg) %>%
  ungroup() %>%
  filter(RegionName %in% names(neighborhoods)) %>%
  pivot_longer(cols = c(zori_avg:hud_avg),
               names_to = "source",
               values_to = "rent") %>%
  mutate(neighborhood = neighborhoods[RegionName])

cleveland_plot <- ggplot(data = cleveland_data,
                         mapping = aes(x = reorder(RegionName, rent), 
                                       y = rent)) +
  geom_line(aes(group = RegionName), size = 1) +
  geom_point(aes(color = source), size = 5) +
  geom_text(data = cleveland_data %>% filter(source == "zori_avg"),
            mapping = aes(color = source,
                          label = paste0("+",scales::percent(x = diff, 
                                                             accuracy = 0.01))),
            size = 3.5,
            hjust = -0.25) +
  geom_text(data = cleveland_data %>% filter(source == "hud_avg"),
            mapping = aes(label = neighborhood),
            size = 3.5,
            hjust = 0,
            nudge_x = 0.5) +
  scale_y_continuous(labels = scales::dollar_format(), limits = c(500, 3500)) +
  coord_flip() +
  labs(title = "Zip Codes With Rent Furthest From Fair Market Rate",
       subtitle = "HUD versus Zillow with Median Income in {current_frame}",
       x = "ZIP",
       y = "Rent Index",
       color = "Source") +
  scale_colour_discrete(labels = c("HUD", "Zillow")) +
  transition_manual(year)

animate(cleveland_plot, fps = 1, nframes = 6)

rm(cleveland_data)
```

```{r fig.width=10, fig.height=11}
### ------------------
### Cleveland Dot Plot 
### Facet wrapped

neighborhoods <-
  list(
    "60045" = "Lake County",
    "60602" = "Downtown",
    "60610" = "Near Northside, Cabrini Green",
    "60126" = "Elmhurst",
    "60642" = "River West",
    "60647" = "Logan Square",
    "60616" = "Chinatown, Armour Square",
    "60525" = "La Grange",
    "60612" = "East Garfield Park",
    "60608" = "Pilsen"
  )

cleveland_data <-
  data %>%
  # can't do functions in label assignments so gotta do it here
  mutate(diff = (zori_avg - hud_avg)/hud_avg) %>%
  ungroup() %>%
  filter(RegionName %in% names(neighborhoods) & year %in% c(2014,2019)) %>%
  pivot_longer(cols = c(zori_avg:hud_avg),
               names_to = "source",
               values_to = "rent") %>%
  mutate(neighborhood = neighborhoods[RegionName])

acs_to_year_mapping <- list(
  "2014" = data_2014,
  "2019" = data_2019
)

# still figuring out the best aesthetic 
cleveland_plot <- function(input_data, yr, acs) {
  plot_data <-
    input_data %>%
    left_join(acs[[as.character(yr)]], by = c("RegionName", "year")) %>%
    filter(year == yr) %>%
    mutate(income_diff = rent/(median_hh_income/12),
           burden = cut(income_diff,
                        breaks=c(0, 0.3, 0.5, 1),
                        labels=c("Low","Medium","High")))
  
  ggplot(data = plot_data,
         mapping = aes(x = RegionName, y = rent)) +
    # line showing the difference between the HUD and Zillow indices
    geom_line(aes(group = RegionName, color = burden), size = 5) +
    # shapes representing either HUD or Zillow index
    geom_point(aes(shape = source), size = 5) +
    # labeling the difference by percentage between indices
    geom_text(
      data = plot_data %>% filter(source == "zori_avg"),
      mapping = aes(
                    label = paste0(
                      "+", scales::percent(x = diff,
                                           accuracy = 0.01)
                    )),
      size = 3.5,
      hjust = -0.25
    ) +
    # labeling the lines with the neighborhood
    geom_text(
      data = plot_data %>% filter(source == "hud_avg"),
      mapping = aes(label = neighborhood),
      size = 3.5,
      hjust = 0,
      nudge_x = 0.5
    ) +
    scale_y_continuous(labels = scales::dollar_format(),
                       limits = c(800, 3200)) +
    coord_flip() +
    labs(
      title = "Zip Codes With Rent Furthest From Fair Market Rate",
      subtitle = "HUD versus Zillow with Average Monthly Median Income",
      x = "ZIP",
      y = "Rent Index",
      shape = "Source",
      color = "Rent Burden"
    ) +
    scale_color_manual(values = c("#00ba38","#83b0fc","#F8766D")) +
    scale_shape_manual(values = c("circle","square"), labels = c("HUD", "Zillow"))
}

library(gridExtra)

grid.arrange(
  cleveland_plot(cleveland_data, 2014, acs_to_year_mapping),
  cleveland_plot(cleveland_data, 2019, acs_to_year_mapping)
)

# rm(cleveland_data)
```

# Limitations

## Missingness in data

# Conclusions

## Findings

HUD FMR jumped significantly from 2017 to 2018. We should look into why that happened. 

## Policy Implications

## Further Questions